---
title: "R Basics"
author: "Randall Pruim"
date: "last-modified"
format: live-html
editor_options: 
  markdown: 
    wrap: 72
webr:
  packages:
    - tidyverse
    - mosaic
    - ggformula
    - readr
    - gt
    - pander
    - reactable
engine: "knitr"
---

{{< include ../_extensions/r-wasm/live/_knitr.qmd >}}


```{webr}
#| label: setup
#| include: false
library(tidyverse)
# library(ggformula)
library(mosaic)
theme_set(theme_bw())
# knitr::opts_chunk$set(exercise.checker = checkr::checkr_tutor)
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  fig.width = 6, fig.height = 2.5)
MI_lead <- read.csv('https://sldr.netlify.app/data/MI_lead.csv')
# file='http://sldr.netlify.app/data/MI_lead.csv')
GDP <- read.csv('http://statistical-modeling-data.netlify.app/GDP.csv')
tutorial_options(exercise.eval = FALSE)
```


## Your Mission

The purpose of this tutorial is to help you start to get familiar with
the way R works, and some basic R commands.

*__Note:__ You won't need to use the RStudio server for this tutorial. But it will
help you use RStudio later.*

<!-- But you may have noticed that *you are not even using R.* At least, you -->
<!-- haven't logged in to the RStudio server. -->

This tutorial environment lets you read some helpful information, then
practice writing and running your own R code *right inside the tutorial!*
Here's hoping it provides a nice, gentle introduction in a controlled environment!

<!-- Tutorials can also serve as helpful reference guides later on, if you -->
<!-- need reminders or more practice. We will keep using them to learn and -->
<!-- practice R throughout the semester, and links to all our tutorials will -->
<!-- be on the course Moodle page ("R Tutorials" section). -->


## Warm Up: Basic arithmetic

R does much more than this, but you can also use it do basic arithmetic.
Hit the run button below to see how it works.

```{webr, arithmetic, exercise = TRUE}
15 + 27
```

#### Your turn: Experiment a little

Edit the code below to compute some other things. Try subtraction, multiplication, division, etc.

```{webr}
#| exercise: arithmetic-your-turn
1 + 1
```

:::{.hint exercise="arithmetic-your-turn"}
**Hint:** Here are a few examples

```r
10 - 2
5 * 5
20 / 3
5^2
```
:::

## Communicating with R

You will do most of your work in R with *code* or *commands*. Instead of
pointing and clicking, you will type one or more lines of code, which R
will *execute* (doing the work you have asked it to do). Then, R will
return the results of whatever operation you asked it to do -- perhaps a 
plot, or some text.
Sometimes executing code has no visible effect 
(no plot or text output is produced), but
instead some object is created and stored in R's *environment* for later
use.

### Two Key Questions

To get R (or any software) to do something for you, there are two
important questions you must be able to answer. Before continuing, think
about what those questions might be.

### The Questions

To get R (or any software) to do a job for you, there are two important
questions you must be able to answer:

#### 1. What do you want the computer to do?

#### 2. What must the computer know in order to do that?

You should always try to answer those two questions before trying 
to figure out what to type.

### Providing R with the information it needs

R *functions* provide R with the answer to the first question: what do
you want the computer to do?

Most functions in R have short, but descriptive names that describe what
they do. For example, R has some functions to do basic mathematical
operations: 

* the function `sqrt()` computes the square root of a number,
* the function `round()` rounds a number (by default, it rounds to the
nearest integer).

or read in data

* the function `read.csv()` reads in data in comma separated value format (a very
common data format that works in many types of software, including Excel.)

But just giving R a function is not enough: you also need to answer the
second question (what information does R need to do the job?). For
example, if you want to use the function `round()`, you also need to
provide R with the number you want to round! (And possibly also the number
of decimal places you want in the result.)

We will provide answers to our two questions by filling in the boxes of
a basic template:

<center>

<h2>

<strong>[function]{.boxed} ( <span class="boxed"> info1, 
info2, ...</span> ) </strong>

</h2>

</center>

The `...` indicates that there may be some additional *arguments*
(input information we could provide to R) we could add eventually. Some
functions need only one input, but if a function takes more than one
argument, they are separated by commas.

### Using simple functions 

Let's practice what you just learned, trying
out the `sqrt()` and `round()` functions.

Edit the code below to compute the square root of 64:

```{webr}
#| exercese: find-sqrt
function(information_R_needs)
```

:::{.hint exercise="find-sqrt"}
:::{.callout-note collapse="true"}
### Hint
```r
sqrt(information_R_needs)
```
:::
:::

:::{.solution exercise="find-sqrt"}

:::{.callout-note collapse="true"}
#### Solution

```r
sqrt(64)
```
:::
:::

Now try computing the square root of 44, then rounding it to the nearest
integer:

```{webr}
#| exercise: round-sqrt
function1(information_R_needs)
function2(information_R_needs)
```

:::{.hint exercise="round-sqrt"}

::::{.callout-note collapse=true}
#### Hint 1

```r
sqrt(information_R_needs)
function2(information_R_needs)
```
::::

::::{.callout-note collapse=true}
#### Hint 2

```{
sqrt(44)
function2(information_R_needs)
```
::::

::::{.callout-note collapse=true}
#### Hint 3

```{webr round-sqrt-hint-3}
sqrt(44)
round(information_R_needs)
```

::::

::::{.callout-note collapse=true}
#### Hint 4

```r
sqrt(44)
round(sqrt(44)) # yes, you can nest functions. sometimes (not always) a good idea.
```

::::
:::

### Storing information in R: variables

In the last section, you computed the square root of 44 and then rounded
it, probably like this:

```{webr, echo=TRUE}
sqrt(44)
round(6.63325)
```

But to do that, you probably had to first find the root, make a note of
the result, and then provide that number to the `round` function. What a
pain!

A better alternative, if you are computing a value that you will want to
use later on, is to store it as a named variable in R. In the previous
example, you might want to store the square root of 44 in a variable
called my_root; then you can provide my_root to the `round` function
without checking the result of the `sqrt` calculation first:

```{webr naming-variable, echo=TRUE}
my_root <- sqrt(44)
round(my_root)
```

Notice that to assign a name to the results of some R code, you use the
symbol `<-`. You can think of it as an *assignment arrow* -- it points
from a result toward a name and assigns the name to the result.

:::{.callout-tip}
I like to read `<-` as "gets": So `myroot <- sqrt(44))` is read as 
"`myroot` gets `sqrt(44)`".
:::

Try editing the code to change the name of the variable from my_root to
something else, then run your new code:

```{webr name-a-variable, exercise=TRUE}
my_root <- sqrt(44)
round(my_root)
```

### What if I have a list of numbers to store?

Sometime you might want to create a variable that contains more than one
number. You can use the function `c` to ***c**oncatenate* or ***c**ombine* several numbers
into a list:

```{webr}
my_fave_numbers <- c(4, 44, 16)
my_fave_numbers
```

(First we stored the list of numbers, calling it my_fave_numbers; then
we printed the results to the screen by simply typing the variable name
my_fave_numbers).

Try making a list of your three favorite numbers, then using the
function `sum` to add them all up:

```{webr}
#| exercise: c-and-sum

```

:::{.solution exercise="c-and-sum"}
#### Solution

```r
my_numbers <- c(14, 27, 455)
sum(my_numbers)
```

### What about data that are not numeric?

R can work with categorical data as well as numeric data. For example,
we could create a list of words and store it as a variable if we wanted
(feel free to try changing the words if you want):

```{webr}
#| label: cat-var
my_words <- c('RStudio', 'is', 'awesome')
my_words
```

### What if I have a LOT more data to store?

`c` works great for creating small lists of just a few values, but it is
not a good way to enter or store large data sets. 

In R, the most common way to store data is in an object called a 
**data frame**. 
The next sections will get you started using data frames in R.

## How should data tables be organized for statistical analysis?

*This section is about **rectangular** data -- data that can be stored in
a rectangular grid of rows and columns, like in an Excel spreadsheet.
This is by the far the most common sort of data used in business and in statistical analysis, but it is not the only kind of data.*

A comprehensive guide to good practices for formatting data tables is
available at <http://kbroman.org/dataorg/>.

A few key points to keep in mind:

-   Organize the table so there is **one column for every variable** 
    (piece of information being recorded), and
    **one row for every observation** (person/place/thing for which data
    were collected).
-   This data table is for the computer to read, not for humans! So
    eliminate formatting designed to make it "pretty" (colors, shading,
    fonts...)
-   Use short, simple variable names that do not contain any spaces or
    special characters (like ?, \$, %, \_, etc.)
-   Use informative variable values rather than arbitrary numeric codes.
    For example, a variable Color should have values 'red', 'white', and
    'blue' rather than 1, 2, and 3.

You will have chances to practice making your own data files and
importing them into R outside this tutorial.

## Using built-in datasets in R

R has a number of built-in datasets that are accessible to you as soon
as you start RStudio.

In addition to the datasets that are included with base R, there are
add-on *packages* for R that contain additional software tools and
sometimes datasets.

To use datasets contained in a package, you have to load the package by
running the command:

```{r, library-example, eval=FALSE}
library(packagename) 
```

The command `require()` does basically the same thing as `library()`.
You can use whichever one you prefer.

```{r require-example, eval=FALSE}
require(packagename)
```

### Example of loading a package

For example, we will practice looking at a dataset from the package
`mosaic`.

Before we can access the data, we have to load the package.

Edit the code below to load the `mosaic` package and then click "run
code" to run it.

**(Nothing obvious will happen...this command just gives R permission to
access the package, without asking it to actually DO anything.)**

```{webr}
#| exercise: require-package
library(______)
```

:::{.hint exercise="require-package"}
**Hint:** You want to load the `mosaic` package.
:::

### Viewing a dataset

Loading the mosaic package also loads dataset called `HELPrct`.

If you just enter the dataset name (`HELPrct`) as a command, R will
print some of the dataset out to the screen:  Give it a try:

```{webr}
#| label: show-HELPrct
HELPrct
```

For a large data frame, this isn't the most useful thing. We'd like to
be able to get some more useful information about the data set.

### Gathering information about a dataset

There are a few functions that make it easier to take a quick look at a
dataset:

-   `head()` prints out the first few rows of the dataset.
-   `glimpse()` gives an overview of the dataset
-   `names()` prints out the names of the variables (columns) in the
    dataset
-   `summary()` prints out a summary of all the variables (columns) in the
    dataset
-   `nrow()` reports the number of rows (observations or cases) in the
    dataset
-   `ncol()` reports the number of columns (variables) in the dataset

Try applying each of these functions to the `HELPrct` data and see what
the output looks like each time:

```{webr}
#| exercise: check-out-data
HELPrct
```

:::{.hint exercise="check-out-data"}

Here are a couple examples.

```r
head(HELPrct)
glimpse(HELPrct)
```
:::


### Getting more help

You can get help related to R function, and built-in R datasets, using a
special function: `?`. Just type `?` followed by the name of the function
or dataset you want help on:

```{webr}
#| exercise: get-help

```

:::{.hint exercise="get-help"}
**Hint:** Try `?HELPrct`.

::::{.callout-note}
The help output looks much nicer inside RStudio than it does inside a tutorial.
::::
:::


## Reading in data from a file

For this class, you will often be asked to analyze data that is stored
in files that are available online -- usually in csv format. It's simple
to read them into R. For example, we can read in the file MI_lead.csv,
which is stored at <https://sldr.netlify.app/data/MI_lead.csv>:

```{webr}
#| label: read-lead-data
MI_lead <- read.csv('https://sldr.netlify.app/data/MI_lead.csv')
```

### Some common mistakes when reading files

The code below contains a couple of the **most common mistakes**
students make when they try to read in a data file. See if you can find
and correct them.

The code below is supposed to read in some GDP data from the file
<http://statistical-modeling-data.netlify.app/GDP.csv>. But there are 
some problems with the code that you need to fix. See if you can 
figure out how to fix it.  Refer to the hints if you have trouble.

One you have the data loaded correctly, check to see how many rows and columns it has.

```{webr}
#| example: gdp-mistakes
read.csv(http://statistical-modeling-data.netlify.app/GPD)
```

:::{.hint exercise="gdp-mistakes"}
::::{.callout-note collapse=true}
#### Hints

1. Be sure to type the file name or url correctly.

2. Don't forget the .csv at the end of the file name

3. The name needs to be in quotation marks.

4. Did you save the data? Did you choose a good name?
::::
:::


:::{.solution exercise="gdp-mistakes"}
::::{.callout-note collapse=true}
#### Solution

```{webr gdp-mistakes-hint-6}
# load and save the data
GDP <- read.csv('http://statistical-modeling-data.netlify.app/GDP.csv')

# glimpse() shows the number of rows and columns and also shows us the 
# first few values of each variable. You could also use `nrow()` and `ncol()` 
# to find the number of rows and columns.

glimpse(GDP)
```
::::
:::


#### Your turn: quick review

Now that we have loaded the MI lead data set,
use one of the functions you learned earlier to find out how many
variables are in the MI_lead dataset, and what their names are.

```{webr}
#| exercise: MI_lead-info

```

:::{.hint exercise="MI_lead-info"}
**Hint:** Try `glimpse()` or `names()`.
:::

### What about local files?

The same function, `read.csv()`, can be used to read in a local file.
You just need to change the input to `read.csv()` -- instead of a URL,
you provide a path and filename (in quotes). For example, we might change

```{r, eval = FALSE}
read.csv('https://sldr.netlify.app/data/MI_lead.csv')
```

to 

```{r, eval = FALSE}
read.csv('C:\\Data\\MI_lead.csv')
```

:::{.callout-caution}
If your operating system uses backslashes (PCs mostly), you will need to type them each **twice**
as in the example above.  Other operating systems (Macs and Linux) use forward slashes (`/`), which
do not need to be typed twice.

The RStudio server is a Linux machine, so it uses forward slashes (`/`).  It also provides 
a way for you to navigate to the file and then it shows you the code required to load that file.
We'll see this when we are working in RStudio.
:::

We won't do an example using a local file in this tutorial because there isn't 
a way to work with local files within a tutorial environment, but you can
practice it once you are working independently in RStudio.

If you are working on [r.cs.calvin.edu](https://r.cs.calvin.edu), 
you will have to *upload* files to your cloud space on the server before you 
can read them in (RStudio on the server cannot access files on your computer's
hard drive).

### Side note: named input arguments

The input argument we provided to R is the URL (in quotes -- either
single or double quotes are fine). The *name* of this argument is `file`, and
there are two ways we can provide the information to R -- 
with the name:

```{webr}
MI_lead <- read.csv(file = 'https://sldr.netlify.app/data/MI_lead.csv')
```

or without the name:

```{webr}
MI_lead <- read.csv('https://sldr.netlify.app/data/MI_lead.csv')
```

If a function only has one argument, it doesn't really matter which way 
we do it.

However, if a function has *more than just one* input argument, then R needs
to know which thing is which, so we must either 

* use names, or
* get things in the order R requires.

In fact, we can mix and match.  Many R functions have one or two
*required* pieces of information that come first in order followed by
many *optional* pieces of information. 
Those optional pieces of information
come in some particular order, but it is easier to just provide the name
and not worry about the order.  That also means we only need to specify the
ones we are interested in changing.

:::{.callout-tip}
Typically we will not use names for first argument or two but use names for
all the rest.
:::

Here is an example.  Let's suppose that you receive a data file from Germany
where a comma is used as the decimal point. That is a problem for a regular
CSV file because the comma is used as the separator between data values. To
get around this, a semicolon has been used as the delimiter.  How do we 
read the resulting file?

If you look at the help for `read.csv()`, you will see that there are a number
of arguments including `sep` (the column-separator) and `dec` (the decimal
point).  So we can read this file using

```{webr}
url <- 'https://calvin-mgmt535.netlify.app/data/german.dat'

german_data <- read.csv(url, sep = ";", dec = ",")
```

As it turns out `sep` is the third argument and `dec` is the fifth. But if 
we use names, we don't need to worry about that or to specify the second and 
fourth arguments.

:::{.callout-tip}
URLs are often rather long. It can be handy to save the URL as a variable (named `url` in the example
above) and the to use that variable in place of the long URL.  This is especially handy if you need
to use the URL more than once.  But it is also useful just to keep the lines shorter and more 
readable.
:::

#### Your turn

Edit the code below so that it rounds to 1 decimal place.  Do this by 
specifying an additional argument: `digits = 1`:

```{webr}
#| exercise: round-arguments
round(1234.567)
```

:::{.solution exercise="round-arguments"}

::::{.callout-note collapse=true}
#### Solution
```r
round(1234.567, digits = 1)
```
::::
:::

<!-- It is good to get in the habit of providing names for the inputs so you don't  -->
<!-- have to worry about the order. If you -->
<!-- provide names, then the order in which you list the inputs doesn't -->
<!-- matter; without names, **the order matters** and you have to use ? to -->
<!-- figure out what order R expects! -->

## Some Bonus Topics

### Using the pipe (`|>`)

If you have already done some plotting with `{ggformula}`, 
you have probably seen the pipe (`|>`) before.
The pipe lets us rewrite

```{r, eval=FALSE}
f(x, ...)
```

as 

```{r, eval = FALSE}
x |> f(...)
```

In other words, whatever is on the left side of `|>` becomes the first argument of the 
function that follows.

This is especially handy when we want to do several things in a row.  Compare

```{r, eval = FALSE}
h(g(f(x, a), b), c)
```

with

```{r, eval = FALSE}
x |> f(a) |> g(b) |> h(c)
```

The pipe makes it easier to tell

* the order of operations: f, then g, then h
* which arguments belong to which operations: a goes with f, b with g, c with h

#### Give it a try

Rewrite each of these commands using the pipe

```{webr}
#| exercise: pipe
glimpse(MI_lead)
round(sqrt(80), digits = 2)
```

:::{.solution exercise="pipe"}
::::{.callout-note collapse=true}
#### Solution

```r
MI_lead |> glimpse()
80 |> sqrt() |> round(digits = 2)
```
::::
:::

:::{.callout-tip}
It can be handy to read the pipe (`|>`) as **then**.  For example: 
"80, **then** take the square root, **then** round to 2 decimal places."
:::

### Renaming variables in a dataset

This is an advanced topic, so don't worry if it seems complicated; for
now, it's just nice to realize some of the power R has to clean up and
reorganize data.

What if we didn't like the names of the `MI_lead` variables? For
example, a new user of the dataset might not know that that ELL stands
for "elevated lead levels" and that ELL2005 gives the *proportion* of
tested kids who had elevated lead levels in the year 2005.

If we wanted to use a clearer (though longer) variable name, we might
prefer `prop_elevated_lead_2005` instead of `ELL2005` -- more letters to
type, but a bit easier to understand for someone unfamiliar with the data.
How can we tell R we want to rename a variable?

We use the code:

```{webr}
#| label: rename-MI_lead
MI_lead2 <- 
  MI_lead |>
  rename(prop_elevated_lead_2005 = ELL2005)

glimpse(MI_lead2)
```

The code above uses some tools you've seen, and some more advanced ones
you haven't seen yet. Translated into words, it tells R:

1. Make a dataset called `MI_lead2` by starting with the dataset `MI_lead`.
2. Then send (`|>`) the `MI_lead` dataset to the function `rename()`.
3. What I want to rename is the variable `ELL2005`. 
Its new name should be `prop_elevated_lead_2005`.

*See...you can already start to make sense of even some pretty
complicated (and useful) code.*

## Check out the data

OK, back to business -- simple functions and datasets in R.

It's your turn to practice now. Use one of the functions you have
learned so far to extract some information about the `MI_lead` dataset.
*Remember, `?` won't work on MI_lead because it's not a built-in R
dataset.*

```{webr}
#| exercise: look-at-MI_lead

```

## Review

What have you learned so far? More than you think!

### Arithmetic in R

Basic arithmetic can be done in R pretty much just like you would do it on a hand-held calculator.
Two big advantages to using are the ability to see/save/edit your arithmetic and the ability
to save the results of your arithmetic for later use.

### Functions in R

You've learned that R code is made up of functions, which are generally
named descriptively according to the job they do. Functions have one or
more input arguments, which is where you provide R with all the data and
information it needs to do the job. The syntax for calling a function
uses the template:

<center>

<h2>

<strong>[function]{.boxed} ( <span class="boxed"> information1  ,
 information2  , ...)</strong>

</h2>

</center>



### Variables in R

You've practiced creating variables in R using `c`, and saving
information (or the results of a computation) using the assignment arrow
\<-.

### Datasets in R

You've considered several different ways to get datasets to work with in
R: you can use datasets that are built in to R or R packages, or you can
use `read.csv` to read in data files stored in .csv format. 

:::{.callout-note}
There are 
similar functions for reading data in several other formats as well,
including excel spreadsheets, data formats used in other statistical
software, etc. If your data are stored in a common data format, you can 
probably get them into R pretty easily.
:::
